Computational methods in landscape ecology

### 2024-09-20 Manuscript.qmd
adds a few comments/changes/etc

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/04-sampling.qmd >}}
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/06-genetics.qmd >}}
{{< include Sections/07-spatial-planning.qmd >}}
{{< include Sections/08-discussion.qmd >}}

@@

We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.

>>>>

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
<!-- JN: Max, should this be uncommented? -->
<!-- {{< include Sections/04-sampling.qmd >}} -->
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/06-genetics.qmd >}}
{{< include Sections/07-spatial-planning.qmd >}}
{{< include Sections/08-discussion.qmd >}}

@@

<!-- JN: Are we proactively thankful? I would suggest to uncomment this sentence and reinstate it when the final version of the paper is accepted -->
We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.

### 2024-09-20 Sections/01-introduction.qmd
Update authors
adds a few comments/changes/etc

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them.
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
![Overview of landscape ecology topics for which computational developments are discussed.](Fig-1.png)

>>>>

Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, including human activities [@Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape]. <!--JN: why do you use "but see" here? is the Wiersma2022 paper contradicting Petrovskii2012?-->
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them.
In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.

![Overview of landscape ecology topics for which computational developments are discussed.](Figures/Fig-1.png){width=80%}

### 2024-09-20 Sections/02-spatial-patterns.qmd
feat: sampling section
adds a few comments/changes/etc

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified. 
This includes shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, such as correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
This is in line with several reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency.
These are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
<!-- MH: I moved sections around quite a bit unfortunately -->

Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology? I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
<!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->

@@

The Landscape Mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes.
The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].

@@

<!-- JN: ask T. Stepinski for review of this section later -->

@@

<!-- JN: ask T. Matuszek for review of this section later -->

@@

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.

>>>>

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of such landscape metrics is that they are easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified. 
This includes shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].

To address some limitations, such as correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify core metrics that capture main components of landscape patterns.
More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
This is in line with several reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundaments of landscape patterns, namely amount and adjacency.
These are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory, and later formalized in @Nowosad2019InformationTheory.

Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology? I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
<!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->
<!-- JN: Agree. There are many R tools related to dimensionality reduction, that could be easily connected to landscape metrics, but I would not mention them here. -->

@@

<!-- JN: why the capitalization of Landscape Mosaic? -->

The Landscape Mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes.
The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
<!-- JN: you mentioned that the method has been improved, but there is no reference to the original paper. -->
<!-- JN: also -- it this idea important enough to have its own section? I assume that there are many papers that propose new ways to classify LULC data. -->
Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].

@@

@@

@@

R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures] or *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for the use in landscape ecology.

### 2024-09-20 Sections/06-genetics.qmd
adds a few comments/changes/etc

Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], among others.
Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].

>>>>

Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], among others.
Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., @Brown2017SDMtoolbox20].

### 2024-09-20 Sections/08-discussion.qmd
adds a few comments/changes/etc

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, documentation [@Tenopir2020DataSharing].

Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Nevertheless, the here provided review of the latest computational methods in landscape ecology can hopefully only be a snapshot in time because new approaches will emerge in the future.

>>>>

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
<!-- JN: what are "large satellites"?  -->
This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
<!-- JN: "is entering" or "has entered"? -->
In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].

Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
Up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Nevertheless, the here provided review of the latest computational methods in landscape ecology can only be a snapshot in time because new approaches will emerge in the future.
